---
title: "JamesCitrate"
output: 
 bookdown::html_document2:
   toc: yes
   fig_caption: yes
   number_sections: false
 bookdown::word_document2:
    reference_docx: "M2_Results_template.docx"
editor_options: 
  chunk_output_type: inline
---

**Code for all diagnostic plots is taken from the book "An R companion to applied regression", John Fox Sanford Weisberg**

# Loading data
```{r data, message=FALSE, echo=TRUE, warning=TRUE}
data <-  read.csv("~/Downloads/Citrate exudation.csv") 
data$Prate <- factor(data$Prate, levels = c("1","5","50")) # Turn pyruvate into a treatment factor, not a numerical variable. It is more of a treatment level, rather then many different pyruvate concentrations. 
data$WAT <- factor(data$WAT, levels = c("3","4","5","6") )
data$CO2 <- factor(data$CO2, levels = c("Ambient","Elevated") )
require(plyr)
data$WAT <- revalue(data$WAT, c("3"="WAT3", "4"="WAT4", "5"="WAT5", "6"="WAT6"))

```
  

  
## A quick look at data  
```{r lineplots, fig.width=7, fig.asp=0.5, echo=TRUE,  message=FALSE, warning=TRUE, fig.cap="A quick look at data using package ggpubr."}

library(ggplot2)
library(ggpubr)
library(reshape2)
data.melt <- melt(data, measure.var = "Citrate")

Lineplt1 <- ggline(data = data.melt, x = "WAT", y = "value", color = "Prate", add = c("mean_se"), ylab = "Citrate concentration")

Lineplt2 <- ggline(data = data.melt, x = "WAT", y = "value", linetype  = "CO2", add = c("mean_se"), ylab = "Citrate concentration")

p <- ggarrange(Lineplt1, Lineplt2, nrow = 1, ncol =2)

p 
```

It looks to me as if in WAT5 something happened to citrate in samples treated with pyruvate concentration '1' and '5' but mostly under elevated CO<sub>2</sub>. But was there an additional effect of pyruvate? I.e. was there an interaction? I will test that in the model. 

<br/>

```{r barplots, fig.width=7, fig.asp=0.5, echo=TRUE,  message=FALSE, warning=TRUE, fig.cap="barplot of data from WAT5 only using package ggpubr."}
library(tidyverse)
dataWAT5 <- data %>% filter(WAT == "WAT5")
dataWAT5 <- melt(dataWAT5, measure.var = "Citrate")

barplot <- ggbarplot(data = dataWAT5, x = "Prate", y = "value", fill = "CO2", add = c("mean_se"), position = position_dodge(0.7) , ylab = "Citrate concentration") + ggtitle("Week 5 after treatment")
barplot
```
Looking at the means from WAT5 it seems that the difference between ambient and elevated did not change with different levels of pyruvate.


<br/>
  
# Model and diagnostic plots
- I treated all predictors for citrate as factors - including pyruvate. This made more sense to me. 
- Hence the model is a 'Multiway Anova'
- Model info on pg 160 - 170++ of book 
- Diagnostic plots on pg 31 - 36

## Using the lm function
```{r model0, message=FALSE, echo=TRUE, warning=TRUE, }
modelcitrate <-  lm(Citrate ~ CO2 * Prate, data = data)
modelcitrate
summary(modelcitrate)

## Anova type 1
anova(modelcitrate)

## Anova type 2
car::Anova(modelcitrate, type = 2)

## Anova type 3 
car::Anova(modelcitrate, type = 3)
```

Confirming there was no interaction. 

## Using the aov function
This gives the exact same results as from the lm function
```{r modelaov, message=FALSE, echo=TRUE, warning=TRUE, }
modelcitrateaov <-  aov(Citrate ~ CO2 * Prate, data = data)
summary(modelcitrateaov)

## Anova type 2
car::Anova(modelcitrateaov, type = 2)

## Anova type 3 
car::Anova(modelcitrateaov, type = 3)
```


## You can also use this package to print model results directly into a neat table
```{r table, message=FALSE, echo=TRUE}
library(sjPlot)
tab_model(modelcitrate)
```
  
<br/>  

## Histogram and qqplot of residuals
```{r model1, message=FALSE, echo=TRUE, warning=TRUE, fig.cap="Histogram of studentized residuals from regression"}
require(car)
hist(rstudent(modelcitrate)) # histogram of studentized residuals from regression
shapiro.test(resid(modelcitrate))
```

<br/>
```{r model2, message=FALSE, echo=TRUE, warning=TRUE, fig.cap="Quantile comparison of studentized residuals from regression. The broken lines show a bootstrapped pointwise 95% confidence envelope for the points"}
qqPlot(modelcitrate)  
```

<br/>

## Testing for influencial samples, i.e. outliers
```{r model3, message=FALSE, echo=TRUE, warning=TRUE, fig.cap="Index plots of Cook's distance and hat values, from the regression", fig.width=7, fig.asp=0.9}
influenceIndexPlot(modelcitrate , vars = c("Cook", "hat"))
```
This looks o.k to me. I learned that Cook's distance > 1 is bad. 


<br/>
```{r model4, message=FALSE, echo=FALSE, warning=TRUE, fig.cap="Component-plus-residdual plots"}
#crPlots(modelcitrate)
```


<br/>  

## Checking for independance of residuals
```{r residuals, fig.width= 8, fig.asp=0.75, message=FALSE, echo=TRUE, warning=TRUE, fig.cap="Residual plot"}
residualPlots(modelcitrate)
```

```{r model5, message=FALSE, echo=TRUE, warning=TRUE, fig.cap="Spread-level plot of studentized residuals from the model"}
spreadLevelPlot(modelcitrate )
ncvTest(modelcitrate)
```

It looks to me as if the residuals are still somewhat dependent on the citrate concentrations. Perhaps the model does not explain the treatments well enough. 


# Linear mixed effect models  
- Build beyond optimal model, meaning a model with all possible explanatory variables and as many interactions as possible  
-  Use AIC or BIC with REML estimation to compare models with the same fixed but different random structures and to derive the best random structure (smallest AIC or BIC value)  
- Use t or F statistic with REML estimation to find the best fixed structure, except you have a model with nested fixed effects, then apply the ML estimation  
- I used the Kenward_Roger approximations as recommended by Halekoh and Höjsgaard, 2014, implemented by Anova test = "F" in the car pck  
- Present the results of the final model with REML estimation

Reference:  

- Zuur, A. F. leno, E. N, Walker, NJ, Saveliev, AA & Smith, G M. 2009: Mixed effects models and extensions in ecology with R.  
- https://www.westernsydney.edu.au/__data/assets/pdf_file/0004/947740/Rmanual_mixedeffects_20160914.pdf  
- http://www.bodowinter.com/tutorial/bw_LME_tutorial2.pdf  

```{r lmpackages, cache = TRUE, message=FALSE, echo=TRUE, warning=TRUE }
library(lme4)
library(lmerTest)
library(car)
library(rafalib)
library(visreg)
library(sjPlot)
library(sciplot)
library(nlme)
library(MASS)
library(ggpubr)
theme_set(theme_bw())
```

### Data overview 
```{r lmegraphs, cache = TRUE,  fig.height=10, fig.asp=1.5, message=FALSE, echo=TRUE, warning=FALSE}

p1.1 <- ggboxplot(data, y = "Citrate", x = "CO2", add = c("jitter"), fill = "CO2")
p1.2 <- ggplot(data, aes(x=(Citrate))) +  geom_histogram(color="black", fill="black") 
p2.1 <- ggboxplot(data, y = "Citrate", x = "Prate", add = c("jitter"), fill = "CO2")
p2.2 <- ggboxplot(data, y = "Citrate", x = "Prate", add = c("jitter"), fill = "gray")
p3.1  <- ggboxplot(data, y = "Citrate", x = "WAT", add = c("jitter"), fill = "CO2")
p3.2 <- ggboxplot(data, y = "Citrate", x = "WAT", add = c("jitter"), fill = "gray", shape = "Prate")

ggarrange(p1.1, p1.2, p2.1, p2.2, p3.1, p3.2, nrow = 3, ncol = 2)
shapiro.test(data$Citrate)
```


### Data overview - Square root transformed
```{r lmegraphsqrt, echo=TRUE, fig.height=10, message=FALSE, warning=FALSE, cache=TRUE}
data$Citrate_sqrt <- sqrt(data$Citrate)
# A quick overview of data
p1.1 <- ggboxplot(data, y = "Citrate_sqrt", x = "CO2", add = c("jitter"), fill = "CO2")
p1.2 <- ggplot(data, aes(x=sqrt(Citrate))) +  geom_histogram(color="black", fill="black") 
p2.1 <- ggboxplot(data, y = "Citrate_sqrt", x = "Prate", add = c("jitter"), fill = "CO2")
p2.2 <- ggboxplot(data, y = "Citrate_sqrt", x = "Prate", add = c("jitter"), fill = "gray")
p3.1  <- ggboxplot(data, y = "Citrate_sqrt", x = "WAT", add = c("jitter"), fill = "CO2")
p3.2 <- ggboxplot(data, y = "Citrate_sqrt", x = "WAT", add = c("jitter"), fill = "gray")

ggarrange(p1.1, p1.2, p2.1, p2.2, p3.1, p3.2, nrow = 3, ncol = 2)
shapiro.test(sqrt(data$Citrate))
```

```{r lmegraphs2, cache = TRUE,  fig.width=7, fig.asp=0.75, message=FALSE, echo=TRUE, warning=FALSE}
ggboxplot(data, y = "Citrate_sqrt", x = "WAT", add = c("jitter"), fill = "CO2", shape = "Prate")
```

### Checking which random structure is most suitable
```{r lmerandom, cache = TRUE}
M <- gls(sqrt(Citrate) ~ CO2 + Prate + CO2:Prate, data = na.omit(data))
M0 <- lme(sqrt(Citrate)  ~  CO2 + Prate + CO2:Prate  , random =~ 1 | WAT, data = na.omit(data))
M1 <- lme(sqrt(Citrate)  ~  CO2 + Prate + CO2:Prate  , random =~ 1 | WAT/Rep, data = na.omit(data))
M2 <- lme(sqrt(Citrate)  ~  CO2 + Prate + CO2:Prate  , random =~ 1 | Rep, data = na.omit(data))
anova(M, M0, M1, M2)
```
Looks like as if the using WAT (M0) as a random factor improves the model

### Next, initial diagnostics of this model
```{r lmediagnostics, echo=TRUE, fig.height=5, message=FALSE, warning=FALSE, cache = TRUE}
data.naomit <- na.omit(data)

E <- resid(M0, type = 'normalized') 
shapiro.test(E)

E <- data.frame(E = E)
Ehist <- ggplot(E, aes(x=(E))) +  geom_histogram(color="black", fill="black") + ggtitle("Residuals Histogram")

F0 <- fitted(M0)
df <- data.frame(F0 = F0, E = E, CO2 = data.naomit$CO2, Prate = data.naomit$Prate)

residscatter <- ggscatter(data= df, x = "F0", y = "E", xlab = "Fitted values", ylab = "Residuals") + ggtitle("Residuals distribution")

CO2scatter <- ggboxplot(data= df, x = "CO2", y = "E", xlab = "CO2 fitted values", ylab = "Residuals", add = ("jitter"))

Pratescatter <- ggboxplot(data= df, x = "Prate", y = "E", xlab = "Prate fitted values", ylab = "Residuals", add = ("jitter"))

ggarrange(Ehist, residscatter, CO2scatter, Pratescatter, nrow = 2, ncol = 2)
```

Looks pretty good. 

### Looking as model summary
```{r lme, echo=TRUE, fig.height=5, message=FALSE, warning=FALSE, cache = TRUE}
summary(M0)
```
